Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-28 paper.tex
Respond to John Feedback results section.

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG). Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL)\@. \john{I don't love ``compatibility with shading languages'' but then we name a single shading language. Instead, maybe something like ``Our implementation targets portability across modern GPU hardware and software platforms'', that sort of thing?} Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable from a portability perspective, as excessively large work tiles risk reducing occupancy on less capable hardware.

@@

Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{StreamScan}.

@@

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.

@@

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

@@

\subsection{Decoupled Fallback Statistics}
To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution: \john{Below, I advocate that we say, for each of the four bullets, what we expect / what would be a good result / what the hardware actually does in practice. The reader doesn't deeply know what to expect / what is good.}
\begin{itemize}
  \item \textbf{Spins per Workgroup:} How many times does a workgroup make a waiting spin?
  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}? \john{Example for above point: something like ``We hope that even with unfair hardware schedulers, this traversal is short (only a few tiles), but we also want to ensure that DF is robust and delivers good performance for even long traversals.''}
  \item \textbf{Total Fallbacks Initiated:} What is the total number of fallback operations initiated across all workgroups?
  \item \textbf{Total Successful Insertions:} What is the total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
\end{itemize}
Despite differing occupancy rates, the number of waiting spins remains consistent across devices. \john{Is this a small number? Large? Tell don't show.} Likewise, lookback length is stable and remains well below the theoretical occupancy bound. Expressed as a percentage of occupancy, it remains under 10\% for all devices, except for the Intel HD620, where its elevated value is likely an artifact of its low absolute occupancy. Most notably, the devices lacking FPG---Apple M3 and M1 Max---do not exhibit significantly higher successful insertion rates than the others, suggesting that true deadlocking is a rare occurrence.

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins.\label{fig:total_spins}}
  \end{subfigure}

@@

  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
    \caption{Lookback Length.\label{fig:lookback_length}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
    \caption{Throughput vs.\ Baseline.\label{fig:execution_time}}
  \end{subfigure}

  \caption{Simulated Blocking. \john{Please write a good caption here---this is an important caption to write! Probably writing subfigure captions is the right way to go (explain each of them individually).}\label{fig:vertical_images}}
\end{figure*}

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. The finer details of this distinction are discussed next.
\section{Discussion}

@@

\subsection{Limits on Speed-of-Light Performance}
Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling. Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores. The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.

Another potential bottleneck is the speed of atomic operations. On some older GPUs---such as pre-Fermi NVIDIA architectures---atomics do not operate within the cache, resulting in significantly higher latency. Slower atomics lead to delayed tile updates, which in turn increase lookback lengths. While our simulated deadlocks provide an artificial means of stress-testing execution, real deadlocks manifest as an effective \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.

>>>>

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers---most notably Apple's M series and ARM---lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for neear SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring seamless integration with modern graphics APIs. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving SOL performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable from a portability perspective, as excessively large work tiles risk reducing occupancy on less capable hardware.

@@

Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers SOL performance without the work tile size tuning required by \emph{StreamScan}.

@@

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved SOL performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.

@@

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

@@

\subsection{Decoupled Fallback Statistics}
To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:

\begin{itemize} 
  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting? With perfectly fair scheduling, no waiting is necessary. Since no real-world scheduler is perfectly fair, we want minimal waiting. Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.

  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}? Ideally, this traversal should be minimal, but longer traversal may result from unfair scheduling. Empirically, however, our data shows that lookback length remains well below the theoretical $O(o)$ bound---typically under 10\% of occupancy---meaning that most workgroups complete \emph{Chained Propagation} in relatively few steps compared to occupancy.

  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups. Ideally, this number would be zero on devices with FPG; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched. In practice, this count is highly dependent on the fairness of the hardware scheduler. On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.

  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}. Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound. Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler. Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
\end{itemize}

Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, indicating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 30.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 1.477---far below our fallback trigger threshold of 4 spins. This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

@@

  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
    \caption{Average number of work tiles traversed before a workgroup completes \emph{Chained Propagation}. Longer lookback lengths indicate less efficient \emph{Chained Propagation}.\label{fig:lookback_length}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
    \caption{Relative execution time of \emph{DF} under increasing blocking rates compared to an unblocked baseline. Decreasing throughput reflects the growing overhead of fallback operations.\label{fig:execution_time}}
  \end{subfigure}

  \caption{\emph{Decoupled Fallback} Statistics Across Increasing Simulated Blocking Rates. Each subfigure highlights a key performance metric affected by the artificial omission of work tile updates, illustrating the resilience of \emph{DF} under extreme scheduling conditions.}
  \label{fig:vertical_images}
\end{figure*}

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency
\section{Discussion}

@@

\subsection{Limits on Speed-of-Light Performance}
Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling. Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores. The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.Another potential bottleneck is the speed of atomic operations. On some older GPUs atomics do not operate within the cache, resulting in significantly higher latency. Slower atomics lead to delayed tile updates, which in turn increase lookback lengths.
